---
# Title: "Reproducible Research: Peer Assessment 1"

### Part 0 - Verify prerequisities: Formatting, required libraries, data
###### Required libraries  
```{r_lib_summary, echo=TRUE}
# Summarizing & reshaping data
library(plyr)         
library(dplyr)   
library(reshape2)

# Plotting data
library(ggplot2)

# Handling dates
library(lubridate)

# Prettyprinting
library(xtable)
library(knitr)


```

###### Required structure
   + \\ -> the directory that contains this file, PA1_template.Rmd
   + \\data -> Activity input data "activity.csv"
   + \\figure -> Figures created by Knitr. Will be created if it doesn't exist. 

```{r global_knitr_options}
# Set global knitr options to ensure that figures are stored in the _figure_ subdirectory
opts_chunk$set(fig.path='figure\\')  

```


### Part 1 - Load and preprocess the data
_Eval 1: Does the submission show code for reading in the dataet and/or processing the data?_    
Yes, this program uses _read.csv_ to read in the dataset, and lubridate's _ymd_ to translate the "Date" string into a time format. 

```{r load_data, echo=TRUE}

# Check to make sure the file exists
input_data <- "data/activity.csv"
ifelse(file.exists("data/activity.csv"), "File exists", stop("File doesn't exist"))

# ... and if it does, read it in
df_activity <- read.csv('data/activity.csv',header=TRUE, as.is=TRUE)
df_activity$date <- ymd(df_activity$date)
```

### Part 2 - What is the mean total of steps taken per day?

_Eval 2: Does the submission contain a histogram of the total number of steps taken each day?_    
Yes, see below.

- Step 1. The activity dataset is summarized by date, calculating the total, mean, and median steps per day.    
- Step 2. The summarized dataset is used to generate a histogram of the total number of steps taken each day. 


```{r hist_total_steps_per_day}
activity_summary_by_day <- summarise(group_by(df_activity,date), sum(steps), mean(steps), median(steps))
names(activity_summary_by_day) <- c("Date", "TotalStepsPerDay", "MeanStepsPerDay", "MedianStepsPerDay")
ggplot(data = activity_summary_by_day, aes(x=Date, y=TotalStepsPerDay), na.rm = TRUE) + geom_bar(colour="black", stat="identity") +  labs(x = "Date", y = "Total number of steps per day") + theme_bw(base_size = 12)

```

_Eval 3: Are both the mean and median number of steps taken each day reported?_  
Yes, see below.

```{r mean_median_table, results="asis"}
kable(activity_summary_by_day)
``` 

### Part 3 - What is the average daily activity pattern?

_Eval 4: Is there a time series plot of the average number of steps taken (averages across all days) versus the 5-minute intervals?_     
Yes, see below.

```{r time_series_plot}

avg_daily_activity_by_interval <- summarise(group_by(df_activity,interval), mean(steps, na.rm=TRUE))
colnames(avg_daily_activity_by_interval) <- c("Interval", "MeanStepsByInterval")
ggplot(data=avg_daily_activity_by_interval, aes(x=Interval, y=MeanStepsByInterval), na.rm = TRUE) + geom_line(colour="black", stat="identity") +  labs(x = "5 minute interval", y = "Mean number of steps taken across all days") + theme_bw(base_size = 12)
```

_Eval 5: Does the report give the 5-minute interval that, on average, contains the maximum number of steps?_     
Yes, see below after the embedded r code. 

```{r max_mean_steps}
# First find the max mean steps
max_mean_steps_by_interval <- max(avg_daily_activity_by_interval$MeanStepsByInterval)
# .. then find the interval associated with the max mean steps
df_max_interval <- avg_daily_activity_by_interval[avg_daily_activity_by_interval$MeanStepsByInterval == max_mean_steps_by_interval,]
# .. and cast it to an integer, since intervals are integers
max_interval <- as.numeric(df_max_interval[1,1])
```
The 5 minute interval with the maximum number of steps: `r max_interval`.

### Part 4 - Imputing missing variables

_Eval 6: Does the report describe and show with code a strategy for imputing missing data?_ 

Yes.
  
    
Description of strategy for imputing missing data:       
1. What was done: Replace missing step values for intervals with the _mean of step values over all invervals_ by day over the entire data set.         
2. Why: Use mean over intervals, not days, to avoid imputing a positive number of steps in intervals where no activity usually occurs.         
   
   
Show with code a strategy for imputing missing data.  
  - See below.   
        
```{r count_missing}
complete_count <- sum(complete.cases(df_activity)) # works because FALSE/missing is counted as 0, TRUE is counted as 1
missing_count <- length(df_activity$steps) - complete_count
```

Note: there are _`r missing_count`_  rows with missing observations.   

```{r impute_data}

# Use dplyr so get a full vector to match the one to be replaced
df_activity_group_step_means_by_interval <- df_activity %>% group_by(interval) %>% mutate(mean(steps, na.rm=TRUE))
names(df_activity_group_step_means_by_interval) <- c("steps", "date", "interval", "mean_steps")

# Now replace all the NAs in Steps with the corresponding means across interval
#  ... invervals not day because some days son't have any observations at all
#  ... means not medians because some intervals have median 0
# First find the NAs in the Steps string
steps_imputed <- df_activity$steps
steps_NA_idx <- which(is.na(steps_imputed)==TRUE)

# get the means to replace the NAs with
activity_means <- df_activity_group_step_means_by_interval$mean_steps

# ... replace them
steps_imputed[steps_NA_idx] <- activity_means[steps_NA_idx]

# ... add finally them back into our original dataset
df_activity <- cbind(df_activity,steps_imputed )
```

_Eval 8. Does the report contain a histogram of the total number of steps taken each day after missing values were imputed?_    

Yes, see below.

```{r Redo_Histogram}

# get means and medians of the steps, both imputed "steps_imputed" and without the imputed values ("steps")
activity_summary_by_day_imputed <- summarise(group_by(df_activity,date), sum(steps_imputed), mean(steps_imputed), median(steps_imputed), sum(steps), mean(steps), median(steps))

# rename the columns to make them easier to deal with
names(activity_summary_by_day_imputed) <- c("Date", "TotalStepsPerDayImputed", "MeanStepsPerDayImputed", "MedianStepsPerDayImputed", "TotalStepsPerDay", "MeanStepsPerDay", "MedianStepsPerDay")

# get the plot
ggplot(data = activity_summary_by_day_imputed, aes(x=Date, y=TotalStepsPerDayImputed), na.rm = TRUE) + geom_bar(colour="black", stat="identity") +  labs(x = "Date", y = "Total number of steps per day (imputed)") + theme_bw(base_size = 12)

```

Calculate and report the mean and median total number of steps taken per day (using the cleaned up data).

```{r report_steps_imputed, results="asis"}
kable(activity_summary_by_day_imputed)
``` 

Is there a difference of the total daily number of steps?     

- Not on days in which there were observations, because each day either had all observations or no observations.  

### Part 5. Differences between weekdays and weekends   

- Create a new factor variable in the data set two levels - weekday and weekend.  

```{r WeekdayWeekend}

# Find the day of the week corresponding to the date in yyyy-mm-dd
df_date_weekday_factor <- as.factor(weekdays(df_activity$date))

# Translate day of the week to weekday/weekend, and add it to the df_activity data frame
df_date_week <- ifelse(df_date_weekday_factor=="Saturday"|df_date_weekday_factor == "Sunday", "Weekend", "Weekday")
df_activity <- cbind(df_activity, df_date_weekday_factor, df_date_week)

```

- Make a panel plot of the 5 minute interval and the avg number of steps taken, averaged across all weekday days or weekend days. 

_Eval 9. Does the report contain a histogram of the total number of steps taken each day after missing values were imputed?_   

Yes, see below. 
```{r imputed}

# Get the mean of the imputed steps by weekday/weekend by date by interval
avg_daily_activity_by_interval_imputed <- summarise(group_by(df_activity,df_date_week, date, interval), mean(steps_imputed, na.rm=TRUE))
colnames(avg_daily_activity_by_interval_imputed) <- c("df_date_week", "date", "interval", "mean_steps_imputed")

# ..and generate the two plots
ggplot(data=avg_daily_activity_by_interval_imputed, aes(x=interval, y=mean_steps_imputed), na.rm = TRUE) + geom_line(colour="black", stat="identity") +  labs(x = "5 minute interval", y = "Mean number of steps taken across all days") + theme_bw(base_size = 12) + facet_grid(df_date_week ~ .) 

```
